Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-06 old/old_thing.tex
bounds on estimate arguments, new region of inconsistency

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lemma}
\label{lem:concavity}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, $t=\{x_1, y_1, x_2, y_2, w\}$, and $\tau^*$ be the InvFels tree.
For a fixed $\{\xi_1,\ldots,\xi_q\}$, the objective function
\begin{align*}
\ell_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits &\left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) \right.\\
&\qquad+ \left.\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau^*, t)\right]
\end{align*}
is concave and
\[
\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau^*, t)
\]
is a unique maximum.
\end{lemma}

\begin{proof}
We focus on the two terms separately, showing they are both concave and, thus, their sum is concave.
Call the second term
\[
\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau^*, t).
\]
For fixed $\xi_j$, the partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
\begin{align*}
\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) &= \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_1}(y_1)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_2}(x_2)\\
&\qquad + \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,y_2}(y_2)+\sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,w}(w).
\end{align*}
Due to this additive form, all off-diagonal terms of the Hessian for this function are zero, so we show that the diagonal terms are nonpositive.
Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
\[
\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,x_1}(x_1).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods}, each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, is
\[
h_{j,x_1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1). \label{eq:partial-x-form}
\end{align}
We have the second derivative
\[
\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
As $x_1\in(0,1)$, we need only $0 \le \sum_j c_je_j \le 1$ to imply the diagonal terms of the Hessian are nonpositive.
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the partial log-likelihood has nonpositive diagonal terms and off-diagonal terms equal to zero.

We now focus on the divergence term.
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$.%, and thus the Hessian of $\shannonDivergence$ is negative definite.
...
\textbf{
The Hessian as a function of $\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t)$ is negative definite, but is it as a function of $t$?
This seems intuitively obvious, but I don't know how to show it.
How could it not be true?
}
%Let D(p, q) = \sum_j p_j \log q_j
%This is concave in q_j, trivially, but we are interested in concavity in t.
%Because t generates q_j, how can this not be true?
%Thought:
%- if f(x) has a unique maximum on [a,b] and g(x) is concave on [a,b], all we want to show is that f(x)+g(x) has a unique maximum on [a,b]; moreover, f(x) is concave in an open ball around its maximum.
%- note that we have D(p, q(\theta)), where q(\theta) is this sort of complicated function involving Hadamard matrices. however, in the trivial case, D(p, q(\theta)) is concave, so we may be able to use induction?
%- note f(x) + sum_h g(x,h) has a unique solution; what about f(x) + max_h g(x,h)? we still need concavity...
%- the sum of divergence and partial likelihood may be easy to show is concave as partial likelihood clearly is.
%- empirical results show that the gradient condition is accurate, though that might mean we need to show there are no zero crossings of the derivative or something like that
%Oblique strategies:
%- the condition of y_2 \ge y_1 and x_2 \ge x_1 may play a role
%- we could also find x,y where derivative has no solutions, so fn is always increasing, though this might require some more argument
%- we could show (if it's true) that going from fidelities to the Hadamard representation results in a concave function in t, and from there we know the divergence is concave in t.
%   this seems hard but may be our best bet; proceed by induction seems reasonable, but, again, hard.
%   it's true in the "trivial" n=0 case, which encourages me.
%   moreover, the fidelities generate the probabilities, and the objective is concave in the probabilities, which is more encouragement.
\end{proof}

In words, this means that, if we have a single maximal ancestral state for each site split, we have a unique maximum.

\begin{lemma}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$ and $\emptyset$ be the maximal ancestral state for all site splits.
There exists a set of $0 < x^*, y^* < 1$ such that $\{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2, 1, 1\}$ is the solution to maximizing \eqref{eq:profile_likelihood}.
\end{lemma}

\begin{proof}
Assume $w=1$ and $y_2=1$.
Then, there is a closed form solution maximizing the joint optimization procedure, as follows.
We simplify the generating probabilities as in Table~\ref{tab:gen-sitepatprob-restricted}.
We see that, on simplifying, for each site split the generating probability is the same functional form as the likelihood.
Thus, the joint optimization procedure has a unique, closed form solution solving
\begin{align*}
&\arg\max_{x_1,y_y,x_2} \ (p_{\emptyset} + p_2 + p_3 + p_{23}) \log(1+x_1) + (p_1 + p_{123} + p_{12} + p_{13}) \log(1-x_1)\\
               &\qquad + (p_{\emptyset} + p_1 + p_3 + p_{13}) \log(1+y_1) + (p_2 + p_{123} + p_{12} + p_{23}) \log(1-y_1)\\
               &\qquad + (p_{\emptyset} + p_1 + p_2 + p_{12}) \log(1+x_2) + (p_3 + p_{123} + p_{13} + p_{23}) \log(1-x_2)\\
&\propto\arg\max_{x_1,y_y,x_2} \ (4+4x*(y^*)^2) \log(1+x_1) + (4-4x*(y^*)^2) \log(1-x_1)\\
               &\qquad + (4+4(y^*)^2) \log(1+y_1) + (4-4(y^*)^2) \log(1-y_1)\\
               &\qquad + (4+4x*(y^*)^2) \log(1+x_2) + (4-4x*(y^*)^2) \log(1-x_2).
\end{align*}
We have estimates
\[
\{\hat{x}_1,\hat{y}_1,\hat{x}_2\} = \{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2\}
\]
and the gradient is equal to zero at this point.

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,1,1\})$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$(1+x_1)(1+y_1)(1+x_2)$\\
    $\{1\}$     & $p_{1}$       &$(1-x_1)(1+y_1)(1+x_2)$\\
    $\{2\}$     & $p_{2}$       &$(1+x_1)(1-y_1)(1+x_2)$\\
    $\{3\}$     & $p_{3}$       &$(1+x_1)(1+y_1)(1-x_2)$\\
    $\{1,2,3\}$ & $p_{123}$   &$(1-x_1)(1-y_1)(1-x_2)$\\
    $\{1,2\}$   & $p_{12}$     &$(1-x_1)(1-y_1)(1+x_2)$\\
    $\{1,3\}$   & $p_{13}$     &$(1-x_1)(1+y_1)(1-x_2)$\\
    $\{2,3\}$   & $p_{23}$     &$(1+x_1)(1-y_1)(1-x_2)$\\
    \hline
\end{tabular}
\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,1,1\}$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:gen-sitepatprob-restricted}
\end{table}

Now we show there are $0 < x^*, y^* < 1$ on which the gradient with respect to $w$ and $y_2$ is increasing on the boundary for all $x_1,y_1,x_2,y_2,w$.
%TODO: is there a unique solution? it's not concave, but i think there may be a unique solution; need to show this... gradient argument?
We use the fact that $x_1=x_2=x$ as there exists a unique solution and the likelihood is invariant to swapping these two variables.
Let
\[
a_{\emptyset} = \frac{(2x)(y_1+y_2)}{(1+x^2)(1+y_1y_2)}
\]
with similar definitions for the remaining site splits.
The objective function proportional to $w$ is
\begin{align*}
\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + a_{\siteSplit_j}w)
\end{align*}
and has gradient with respect to $w$ as
\begin{align*}
\ell_w(t) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{1 + a_{\siteSplit_j}w}.
\end{align*}
Note that, $a_{\emptyset} = -a_{13}$, $a_{2} = -a_{123}$, $p_{2} = p_{123}$, and
\[
a_{1} = a_{3} = a_{12} = a_{23} = 0
\]
so
\begin{align*}
\ell_w(t) = \frac{1}{1+w} + \frac{p_{\emptyset}a_{\emptyset}}{1 + a_{\emptyset}w} - \frac{p_{13}a_{\emptyset}}{1 - a_{\emptyset}w} + \frac{p_{2}a_{2}}{1 + a_{2}w} - \frac{p_{2}a_{2}}{1 - a_{2}w}.
\end{align*}

For $y_2$, we do the same, but now with
\begin{align*}
\ell(y_2) \propto \log(1+y_2) + \sum_j p_{\siteSplit_j} \log(1 + b_{\siteSplit_j}y_2)
\end{align*}
where
\[
b_{\emptyset} = \frac{y_1+x^2y_1+2wx}{1+x^2+2wxy_1}.
\]
The denominator is always positive, allowing the above simplification.
%Since x_1=x_2 at the maximum we can use an argument like 1+x^2-2y_1wx > 1+x^2-2x = (1-x)^2 > 0. We'll need to ignore cases where x=0 (i.e., x^* or y^* are zero)
In Fig.~\ref{fig:positive-gradients}, we see the region where the gradients of both $w$ and $y_2$ are increasing at $\hat{t}=\{x^*(y^*)^2, (y^*)^2, x^*(y^*)^2, 1, 1\}$, and, thus, the objective function is increasing in all positive directions over $(y_2,w)$ on the boundary.
Since, from Lemma~\ref{lem:concavity}, the objective function is concave on our domain, this must be the maximum.

Ok, the feasible set for the solution is actuall not $[0,1]^5$ because of assumptions we've made.
It's actually
\begin{itemize}
\item $x_1 \in [0,x_2]$
\item $y_1 \in [0,y_2]$
\item $x_2 \in [0,1]$
\item $y_2 \in [y^*,1]$
\item $w \in [y^*,1]$
\end{itemize}
Now, it's possible that our objective function, while not concave, is still decreasing in all directions on this set from the maximum.
How can we show this?
Also, do we want to prove the objective is not concave?
We might be able to have better bounds, like
\begin{itemize}
\item $x_1=x_2 \in [0,x^*]$
\item $y_1 \in [0,y^*]$
\item $y_2 \in [y^*,1]$
\item $w \in [y^*,1]$
\end{itemize}
Bound:
\begin{align*}
\ell_w(t) = \frac{1}{1+w} + \frac{p_{\emptyset}a_{\emptyset}}{1 + a_{\emptyset}w} - \frac{p_{13}a_{\emptyset}}{1 - a_{\emptyset}w} + \frac{p_{2}a_{2}}{1 + a_{2}w} - \frac{p_{2}a_{2}}{1 - a_{2}w}.
\end{align*}

\end{proof}

\begin{figure}
\centering
\includegraphics[width=.95\textwidth]{positive-gradients-inkscape}
\caption{
Region where gradients w.r.t. $w$ and $y_2$ are increasing on the boundary.
}
\label{fig:positive-gradients}
\end{figure}
